VISUAL FEATURES OF INSCRIPTIONS. An issue for EDB (and EAGLE)

### 2016-02-09 source/40/Felle, Visual Features of inscriptions (EAGLE2016_Full_Paper).tex
Corrections to Felle

After storing in EDB other 10000 inscriptions, since 2012 to the present day, the percentage of figured inscriptions is
still the same: then, I think that we are able to consider this datum enough sure; so we are able to partially correct
the common idea that using images in written monuments is a recurrent, typical and characteristic feature of almost all
the Early Christian inscriptions.

@@

Carlo Carletti explained this phenomenon as the result of a will to display explicit signs of a religious identity, such
the Chi-Rho monogram is (\citealp[68-72]{carletti_epigrafia_2008}; \citealp[365-366]{felle_judaism_2007}). But we have to underline that the phenomenon is
not exclusive of Christian patrons. We observe more and more recurrent similar ``signs of
identity'' also among inscriptions commissioned by Jews, not only in Rome but also in other contexts in
Late Antiquity world where they were (\citealp[passim]{felle_judaism_2007}; \citealp{pesce_documenti_2016}).

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img001.pdf}
\caption{Use of different images in inscriptions by Christians of Rome between III and IV cent. (chart by A. E. Felle)}
\label{fig:2}
\end{figure}

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img002.jpg}
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img003.png}
\caption{Rome, \emph{coemeterium Hippolyti}, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.12.14) and
edition in ICVR, VII 19820.}
\label{fig:3}
\end{figure}

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img004.png}
  \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img005.png}

\caption{Rome, \emph{coemeterium Maius}, now in the Vatican Museums. Photo (from Iscrizioni 1997, sch. 3.8.3) and edition in
ICVR, VIII 22407.}
\label{fig:4}
\end{figure}

@@

\noindent The issue of the ``illustrated inscriptions'' brings us to consider more broadly the visual
features of inscriptions overall. The ancient epigraphs are conceived not only as texts to be read (very few people
were able to do it) but also - and, maybe, firstly - as objects to be seen. I think that we have to realize this
perspective - common in Christian epigraphy studies after all - to understand the communication power of this
communicative \emph{medium}. Indeed, the topic has been assumed as main theme of the last International Congress of Greek and
Latin Epigraphy in Berlin in 2012 \citep{eck_vol_2014}; very recently, a just published volume collects various essays
about this same topic just about the Late Antique, Medieval (both Christian and Islamic) world, where this notion of
the inscriptions as 'written monuments' stands out with strong evidence \citep[see][]{eastmond_viewing_2015}.

In our projects - we have to admit it - the notion that inscriptions are \emph{essentially} texts is still largely
prevailing: but now we can - consequently, we must - to increase our capability to represent, to record and so to
investigate also other, visual features of inscriptions.

@@

\noindent First of all, I think to the positioning of the epigraphs in the contexts for which they were created. About EDB, we
already presented in the EAGLE Conference in Paris in 2014 a first attempt to record and to describe the exact
positioning of the inscriptions still found in their original spot, by sharing data with the Domitilla-Projekt (by the
Österreichische Akademie der Wissenschaften and the Deutsches Archäologisches Institut) focused to the frescoes of the
catacomb of Domitilla along via Ardeatina, the largest one in Rome suburb \citep{zimmermann_epigraphy_2014}. The positive
collaboration between our two projects continued: now we are able to offer to EDB users to view on the updated map of
the catacomb of Domitilla the distribution of the inscriptions still \emph{in situ} in the four levels of the subterranean
cemetery: one can use them as reliable documents to (re-)consider the history of the complex, to confirm or to deny the
ideas about chronology of its excavation and of its frescoes and about the using of the different zones of the
catacomb.

@@

\noindent Materials and shape of the written objects communicate immediately, \emph{to all}, before the inscribed texts. Because now we
can do it, we have to display to the users of our projects the real communicative power of the inscriptions, that
assume much more meaningfulness when we can see them than when we can read their only texts in a library. By pictures
and 3D rendering of the places where the epigraphs were arranged we should be able also to provide virtual images of
inscriptions {}- also lost or incomplete - conceived to be seen and read exactly there. An effective example can be
the dedication by pope Damasus (366-384) to the martyr \emph{Ianuarius} in the catacomb of Praetextatus on the via Appia (Fig.~\ref{fig:6}).

@@

\begin{figure}[!bp]
\centering
 \includegraphics[scale=0.06]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img010.jpg}
\caption{Thessaloniki, city walls. Inscriptions in the masonry of a tower near Eptapyrgion (photo: A. E. Felle).}
\label{fig:7}
\end{figure}

@@

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img011.jpg}
\caption{Rome, basilica of the martyr Agnes on the via Nomentana. Apse mosaic with the image of the martyr with the
“useless” caption  s(an)c(t)a Agnes (photo: A. E. Felle).}
\label{fig:8}
\end{figure}

@@

On the open \emph{codex} is inscribed a focused - but barely readable - quotation from John 10, verses 7 and 9, where Jesus
indicates himself as the gate:

@@

\end{quotation}
\begin{figure}[!bp]
\centering
\begin{minipage}[c]{0.42\textwidth}
%\begin{figure}
%\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img012.jpg}
%\end{figure}
\end{minipage}
\begin{minipage}[c]{0.57\textwidth}
%\begin{figure}
%\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img013.jpg}
%\end{figure}
\end{minipage}
\caption{Istanbul, Hagia Sophia. On the left, the Royal Gates in the narthex. On the right, the particular of the image
of the throne with open inscribed codex above the lintel of the central gate (from Kähler 1967).}
\label{fig:9}
\end{figure}

@@

\begin{quotation}
((crux)) \begin{otherlanguage*}{greek}
εἶπεν ὁ κ(ύριο)ς \textlatin{|} ἐγώ εἰμι \textlatin{|} ἡ θύρα τῶν | προβάτων· \textlatin{|} δι᾿ ἐμοῦ \textlatin{||} ἐάν τις \textlatin{|} εἰσέλθῃ \textlatin{|} εἰσελεύσετ(αι) \textlatin{|} κ(αὶ) ἐξελεύσετ(αι) \textlatin{|} κ(αὶ) νομὴν \textlatin{|} εὑρήσει.
\end{otherlanguage*}

@@

The archeological context and the mirate positioning make tangible, concrete, the sacred words; and the real presence
(not necessarily the readability) of the sacred words give proper and strong sense to their material support and to
entire context, the Royal Gates of the Great Church of Constantinoples \citep[320 and passim]{felle_esporre_2015}.

@@

\noindent Scarce or null readability does not imply low quality of the appearance of writing: rather, the writing appears
intrinsecally meaningful such as visual element of the equipment of a simple or rich funerary monument or of a cultual
building. Then, we have to face the issue of the description of the writing not only from the necessary point of
view of paleography ( we are still waiting for a shared and controlled vocabulary of paleographical definitions), but
also in order to perceive and to understand its non-verbal significance: by its disposition, direction, shape. The
notion of the inscriptions in the Islamic world, where often the letters are also - and maybe firstly - images (they
are used as decorative friezes, architectural ornaments, figures) and their clearity and readability are not considered
as necessary, can help us to evocate this feature of the writing (Fig.~\ref{fig:10}).

\begin{figure}[!bp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img014.jpg}
\caption{Granada, Alhambra. An example of the writing as decorative frieze and architectural ornaments (Photo: A. E.
Felle)}
\label{fig:10}
\end{figure}

@@

In conclusion: we have to consider in digital descriptions of the inscriptions some their``visual
features'' that in our projects - first of all in EDB, of course - are not too considered, although they
are very significant. We need, about encoding these non-verbal features, the same positive results that by Epidoc we
reached in encoding the texts: a hard challenge.

>>>>

After storing in EDB other around 10000 inscriptions, since 2012 to the present day, the percentage of figured inscriptions is
still the same: then, I think that we are able to consider this datum enough sure; so we are able to partially correct
the common idea that using images in written monuments is a recurrent, typical and characteristic feature of almost all
the Early Christian inscriptions.

@@

Carlo Carletti explained this phenomenon as the result of a will to display explicit signs of a religious identity, such
the Chi-Rho monogram is (\citealp[68-72]{carletti_epigrafia_2008}; \citealp[365-366]{felle_judaism_2007}). But we have to underline that the phenomenon is
not exclusive of Christian patrons. We observe more and more recurrent similar ``signs of religious identity'' also among inscriptions commissioned by Jews, not only in Rome but also in other contexts in
Late Antiquity world where they were (\citealp[passim]{felle_judaism_2007}; \citealp{pesce_documenti_2016}).

@@

\begin{figure}[!htbp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img001.pdf}
\caption{Use of different images in inscriptions by Christians of Rome between III and IV cent. (chart by A. E. Felle)}
\label{fig:2}
\end{figure}

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img002.jpg}
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img003.png}
\caption{Rome, \emph{coemeterium Hippolyti}, now in the Vatican Museums. Photo (from \emph{Iscrizioni} 1997, sch. 3.12.14) and
edition in ICVR, VII 19820.}
\label{fig:3}
\end{figure}

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img004.png}
  \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img005.png}

\caption{Rome, \emph{coemeterium Maius}, now in the Vatican Museums. Photo (from \emph{Iscrizioni} 1997, sch. 3.8.3) and edition in
ICVR, VIII 22407.}
\label{fig:4}
\end{figure}

@@

\noindent The issue of the ``illustrated inscriptions'' brings us to consider more broadly the visual
features of inscriptions overall. The ancient epigraphs are conceived not only as texts to be read (very few people
were able to do it) but also - and, maybe, firstly - as objects to be seen. I think that we have to realize this
perspective - common in Christian epigraphy studies, after all - to understand the communication power of this
communicative \emph{medium}. Indeed, the topic has been assumed as main theme of the last International Congress of Greek and
Latin Epigraphy in Berlin in 2012 \citep{eck_vol_2014}; very recently, a just published volume collects various essays
about this same topic just about the Late Antique, Medieval (both Christian and Islamic) world, where this notion of
the inscriptions as 'written monuments' stands out with strong evidence \citep[see][]{eastmond_viewing_2015}.

In our projects - we have to admit it - the notion that inscriptions are \emph{essentially} texts is still largely
prevailing: but now we can --- consequently, \emph{we must} --- to increase our capability to represent, to record and so to
investigate also other, visual features of inscriptions.

@@

\noindent First of all, I think to the positioning of the epigraphs in the contexts for which they were created. About EDB, we
already presented in the EAGLE Conference in Paris in 2014 a first attempt to record and to describe the exact
positioning of the inscriptions still found in their original spot, by sharing data with the Domitilla-Projekt (by the
Österreichi\-sche Akademie der Wissenschaften and the Deutsches Archäologisches Institut) focused to the frescoes of the
catacomb of Domitilla along via Ardeatina, the largest one in Rome suburb \citep{zimmermann_epigraphy_2014}. The positive
collaboration between our two projects continued: now we are able to offer to EDB users to view on the updated map of
the catacomb of Domitilla the distribution of the inscriptions still \emph{in situ} in the four levels of the subterranean
cemetery: one can use them as reliable documents to (re-)consider the history of the complex, to confirm or to deny the
ideas about chronology of its excavation and of its frescoes and about the using of the different zones of the
catacomb.

@@

\noindent Materials and shape of the written objects communicate immediately, \emph{to all}, before the inscribed texts. Because now we
can do it, we have to display to the users of our projects the real communicative power of the inscriptions, that
assume much more meaningfulness when we can see them than when we can read their only texts in a library. By pictures
and 3D rendering of the places where the epigraphs were arranged we should be able also to provide virtual images of
inscriptions {}- also lost or incomplete - conceived to be seen and read exactly there. An effective example can be
the dedication by pope Damasus (366-384) to the martyr \emph{Ianuarius} in the catacomb of Praetextatus along the via Appia (Fig.~\ref{fig:6}).

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img010.jpg}
\caption{Thessaloniki, city walls. Inscriptions in the masonry of the so-called Lapardas Tower near `Eptapyrgion' (Komnenian period, XII cent.) (photo: A. E. Felle).}
\label{fig:7}
\end{figure}

@@

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img011.jpg}
\caption{Rome, basilica of the martyr Agnes on the via Nomentana. Apse mosaic with the image of the martyr with the
``useless'' caption  \emph{s(an)c(t)a Agnes}, age of pope Honorius I, 625/638. (photo: A. E. Felle).}
\label{fig:8}
\end{figure}

@@

\begin{figure}[!hbp]
\centering
\begin{minipage}[c]{0.42\textwidth}
%\begin{figure}
%\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img012.jpg}
%\end{figure}
\end{minipage}
\begin{minipage}[c]{0.57\textwidth}
%\begin{figure}
%\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img013.jpg}
%\end{figure}
\end{minipage}
\caption{Istanbul, Hagia Sophia. On the left, the Royal Gates in the narthex. On the right, the particular of the image
of the throne with open inscribed codex above the lintel of the central gate (from Kähler 1967).}
\label{fig:9}
\end{figure}

\newpage
On the open \emph{codex} is inscribed a focused - but barely readable - quotation from John 10, verses 7 and 9, where Jesus
indicates himself as the gate:

@@

@@

\begin{quotation}
((crux)) \begin{otherlanguage*}{greek}
εἶπεν ὁ κ(ύριο)ς \textlatin{|} ἐγώ εἰμι \textlatin{|} ἡ θύρα τῶν \textlatin{|} προβάτων· \textlatin{|} δι᾿ ἐμοῦ \textlatin{||} ἐάν τις \textlatin{|} εἰσέλθῃ \textlatin{|} εἰσελεύσετ(αι) \textlatin{|} κ(αὶ) ἐξελεύσετ(αι) \textlatin{|} κ(αὶ) νομὴν \textlatin{|} εὑρήσει.
\end{otherlanguage*}

@@

The archeological context and the mirate positioning make tangible, concrete, the sacred words; and the real presence
(not necessarily the readability) of the sacred words give proper and strong sense to their material support and to
entire context, the Royal Gates of the Great Church of Constantinoples \citep[320 and \emph{passim}]{felle_esporre_2015}.

\vspace{2cm}

@@

\noindent Scarce or null readability does not imply low quality of the appearance of writing: rather, the writing appears
intrinsecally meaningful such as visual element of the equipment of a simple or rich funerary monument or of a cultual
building. Then, we have to face the issue of the description of the writing not only from the necessary point of
view of palaeography ( we are still waiting for a shared and controlled vocabulary of paleographical definitions), but
also in order to perceive and to understand its non-verbal significance: by its disposition, direction, shape. The
notion of the inscriptions in the Islamic world --- where often the letters are also and maybe firstly images (they
are used as decorative friezes, architectural ornaments, figures) and their clarity and readability are not considered
as necessary --- can help us to evocate this feature of the writing (Fig.~\ref{fig:10}).

\begin{figure}[!hbp]
\centering
 \includegraphics[width=\columnwidth]{FelleVisualFeaturesofinscriptionsEAGLE2016FullPaper-img014.jpg}
\caption{Granada, Alhambra. An example of the writing as decorative frieze and architectural ornaments (Photo: A. E.
Felle)}
\label{fig:10}
\end{figure}

@@

In conclusion: we have to consider in digital descriptions of the inscriptions some their``visual
features'' that in our projects - first of all in EDB, of course - are not too considered, although they
are very significant. We need, about encoding these non-verbal features, the same positive results that by EpiDoc we
reached in encoding the texts: a hard challenge.

\nocite{Manzella1997Felle}
